### 2017-12-22 paper/jfp-submission/evaluation.tex
minor

  let _ = wwhile ((*@\hlOcaml{\hlSherrloc{f}}@*), 2)
\end{ecode}
%
The student has forgotten to \emph{apply} @f@ at all on line 2,
and just matches it directly against a pair.
This faulty @wwhile@ definition nevertheless typechecks,
and is assumed to be correct by both \ocaml\ and \sherrloc\
which blame the use-site on line 10.
%
\begin{verbatim}
  This expression has type
    int -> int * bool
  but an expression was expected of type
    'a * bool
\end{verbatim}
%
\toolname\ synthesizes a trace that draws the eye to the
true error: the @match@ expression on line 2, and highlights
the conflict in matching a function against a pair pattern.
%
\begin{center}
  \includegraphics[height=135px]{wwhile.png}
\end{center}
%
By highlighting conflicting values, \ie\ the source and sink
of the problem, and not making assumption about function correctness, \toolname\
focusses the user's attention on the piece of code that is
actually relevant to the error.

>>>>

  let _ = wwhile ((*@\hlOcaml{\hlSherrloc{f}}@*), 2)
\end{ecode}
%
The student has forgotten to \emph{apply} @f@ at all on line 2,
and just matches it directly against a pair.
This faulty @wwhile@ definition nevertheless typechecks,
and is assumed to be correct by both \ocaml\ and \sherrloc\
which blame the use-site on line 10.
%
\begin{verbatim}
  This expression has type
    int -> int * bool
  but an expression was expected of type
    'a * bool
\end{verbatim}
%
\toolname\ synthesizes a trace that draws the eye to the
true error: the @match@ expression on line 2, and highlights
the conflict in matching a function against a pair pattern.
%
\begin{center}
  \includegraphics[height=135px]{wwhile.png}
\end{center}
%
By highlighting conflicting values, \ie\ the source and sink
of the problem, and not making assumptions about function correctness, \toolname\
focusses the user's attention on the piece of code that is
actually relevant to the error.

### 2017-12-22 paper/jfp-submission/related.tex
minor

\paragraph{Localizing and Repairing Type Errors}
\label{sec:diagnosis-repair}
% It is well known that
% unification-based type inference procedures can
% produce poor error messages, and in particular, can misidentify the
% \emph{source} of the type error.
%
Many groups have explored techniques to pinpoint the true source
of errors reported by static type checkers.
% unification-based type inference procedures.
%
% \subparagraph{Localization}
The traditional Damas-Milner type inference algorithm~\cite{Damas1982-uw}
reports the first program location where a type mismatch is discovered
(subject to the traversal strategy~\cite{Lee1998-ys}).
%
As a result the error can be reported far away from its
source~\cite{McAdam1998-ub} without enough information to guide the
user.
%
Type-error slicing~\cite{Haack2003-vc,Schilling2011-yf,Rahli2015-tt,Sagonas2013-bf,Gast2004-zd,Neubauer2003-xv}
%treats type inference as a constraint-satisfaction problem and
recognizes this flaw and instead produces a slice of the program
containing \emph{all} program locations that are connected to the type
error.
%
%% \cite{Neubauer2003-xv} present a decidable type system based on
%% discriminative sum types, in which all terms are typeable and type
%% derivations contain all type errors in a program. They then use the
%% typing derivation to slice out the parts of the expression related to
%% each error.
%
Though the program slice must contain the source of the error, it can
suffer from the opposite problem of providing \emph{too much}
information, motivating recent work in ranking the candidate locations.
%
Zhang~\etal~\shortcite{Zhang2014-lv,Zhang2015-yu} present an algorithm for
identifying the most likely culprit using Bayesian reasoning.
%
Pavlinovic~\etal~\shortcite{Pavlinovic2014-mr,Pavlinovic2015-kh}
translate the %error
localization problem to a MaxSMT optimization problem, using
compiler-provided weights to rank the possible sources.
%
Loncaric~\etal~\shortcite{Loncaric2016-uk} improve the scalability of
Pavlinovic~\etal by reusing the existing type checker as
a theory solver in the Nelson-Oppen~\shortcite{Nelson1979-td}
style, thus requiring only a MaxSAT solver.

@@

\paragraph{Improving Error Messages}
%
The content and quality of the error messages themselves has also been
studied extensively.
%
Marceau~\etal~\shortcite{Marceau2011-ok,Marceau2011-cy} study the effectiveness of error
messages in novice environments and present suggestions for improving
their quality and consistency.
%
Hage~\&~Heeren~\shortcite{Hage2006-hc} identify a variety of general heuristics to improve
the quality of type error messages, based on their teaching experience.
%
Heeren~\etal~\shortcite{Heeren2003-db},
Christiansen~\shortcite{Christiansen2014-qc}, and
Serrano~\&~Hage~\shortcite{Serrano2016-oo}
provide methods for library authors to specialize
type errors with domain-specific knowledge.
%
The difference with our work is more pronounced here as we do not
attempt to improve the quality of the error message, instead we search
for a witness to the error and explain it with the resulting execution
trace.
%

@@

A function's type is a theorem about the its behavior.
Thus, \toolname's witnesses can be viewed as \emph{counter-examples},
thereby connecting it to work on using test cases to find
counter-examples prior to starting a proof~\cite{ACL2Testing, Seidel15}.

>>>>

\paragraph{Localizing and Repairing Type Errors}
\label{sec:diagnosis-repair}
% It is well known that
% unification-based type inference procedures can
% produce poor error messages, and in particular, can misidentify the
% \emph{source} of the type error.
%
Many groups have explored techniques to improve the error locations
reported by static type checkers.
% unification-based type inference procedures.
%
% \subparagraph{Localization}
The traditional Damas-Milner type inference algorithm~\cite{Damas1982-uw}
reports the first program location where a type mismatch is discovered
(subject to the traversal strategy~\cite{Lee1998-ys}).
%
As a result the error can be reported far away from its
source~\cite{McAdam1998-ub} without enough information to guide the
user.
%
Type-error slicing~\cite{Haack2003-vc,Schilling2011-yf,Rahli2015-tt,Sagonas2013-bf,Gast2004-zd,Neubauer2003-xv}
%treats type inference as a constraint-satisfaction problem and
recognizes this flaw and instead produces a slice of the program
containing \emph{all} program locations that are connected to the type
error.
%
%% \cite{Neubauer2003-xv} present a decidable type system based on
%% discriminative sum types, in which all terms are typeable and type
%% derivations contain all type errors in a program. They then use the
%% typing derivation to slice out the parts of the expression related to
%% each error.
%
Though the program slice must contain the source of the error, it can
suffer from the opposite problem of providing \emph{too much}
information, motivating recent work in ranking the candidate locations.
%
Zhang~\etal~\shortcite{Zhang2014-lv,Zhang2015-yu} present an algorithm for
identifying the most likely culprit using Bayesian reasoning.
%
Pavlinovic~\etal~\shortcite{Pavlinovic2014-mr,Pavlinovic2015-kh}
translate the %error
localization problem to a MaxSMT optimization problem, using
compiler-provided weights to rank the possible sources.
%
Loncaric~\etal~\shortcite{Loncaric2016-uk} improve the scalability of
Pavlinovic~\etal by reusing the existing type checker as
a theory solver in the Nelson-Oppen~\shortcite{Nelson1979-td}
style, thus requiring only a MaxSAT solver.

@@

\paragraph{Improving Error Messages}
%
The content and quality of the error messages themselves has also been
studied extensively.
%
Marceau~\etal~\shortcite{Marceau2011-ok,Marceau2011-cy} study the
effectiveness of error messages in novice environments and present
suggestions for improving their quality and consistency.
%
Hage~\&~Heeren~\shortcite{Hage2006-hc} identify a variety of general
heuristics to improve the quality of type error messages, based on their
teaching experience.
%
Chargu{\'e}raud~\shortcite{Chargueraud2015-dc} presents a tabular format
for type errors that can provide multiple explanations in a compact form.
%
Heeren~\etal~\shortcite{Heeren2003-db},
Christiansen~\shortcite{Christiansen2014-qc}, and
Serrano~\&~Hage~\shortcite{Serrano2016-oo}
provide methods for library authors to specialize
type errors with domain-specific knowledge.
%
The difference with our work is more pronounced here as we do not
attempt to improve the quality of the error message, instead we search
for a witness to the error and explain it with the resulting execution
trace.
%

@@

A function's type is a theorem about its behavior.
Thus, \toolname's witnesses can be viewed as \emph{counter-examples},
thereby connecting it to work on using test cases to find
counter-examples prior to starting a proof~\cite{ACL2Testing, Seidel15}.

### 2017-12-22 paper/jfp-submission/semantics.tex
minor

\paragraph{Narrowing Types}
The procedure % $\force{v}{t}{\vsu}{\tsu}$,
\[
\forcesym : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
\]%
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
%
\begin{figure}[t]
\[
\begin{array}{lll}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
% \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
% \begin{cases}
%   \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
%   \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\

%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
% \end{cases} \\
\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq % \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    &\ \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}, \\
                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\end{array} \\
  \triple{\stuck}{\vsu}{\tsu} &\ \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}\end{array} \\
  \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} &\ \textbf{if} \begin{array}{l} \tsu' = \unify{\{\thole, t\}}{\tsu}, \\ v = \gen{t}{\tsu'} \end{array}\\
\end{cases}  \\
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''} & \hspace{-43.5mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu},\\ \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'} \end{array} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq \triple{\vleaf{t_1}}{\vsu}{\tsu'} & \hspace{-43.5mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array}\\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} \hspace{-3mm} & \defeq \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'} & \hspace{-43.5mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array} \\
\force{v}{t}{\vsu}{\tsu} & \defeq \triple{\stuck}{\vsu}{\tsu}
\end{array}
\]
\caption{Narrowing values}
\label{fig:narrow}
\end{figure}
%
%While a hole may map to a value that \emph{contains} another hole, \eg a
%lambda or a tree, it may not map \emph{directly} to another hole,
As the value substitution is normalized, in the first case of \forcesym\ we
do not need to \forcesym\ the result of the substitution, the sub-hole
will be narrowed when the context demands it.

@@

\paragraph{The Type of a Value} The \emph{dynamic type}
of a value $v$ is defined as a function $\typeof{v}$ shown
in Figure~\ref{fig:typeof}.
%
The types of primitive values are defined in the natural manner.
%
The types of functions are \emph{approximated}, which is all
that is needed to ensure an application does not get stuck.
%
For example,
$$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$$
instead of $\tint \rightarrow \tint$.
%
The types of (polymorphic) trees are obtained from the labels on their
values, and the types of tuples directly from their values.

@@

In order to convert the semantics into a \emph{procedure} for finding
witnesses, we must address two challenges.
%
First, we must resolve the non-determinism introduced by \gensym.
%
Second, in the presence of higher-order functions and currying,
we must determine how many concrete values to generate to make
execution go wrong (as we cannot rely upon static typing to
provide this information.)

@@

\begin{cor}[Witness Generation]
\label{thm:generation}
  If $\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}
\begin{proof}
  For any function $f$ of multiple arguments, we can define $f'$ as the
  uncurried version of $f$ that takes all of its arguments as a single
  nested pair, and then apply Theorem~\ref{thm:soundness} to $f'$.
\end{proof}

>>>>

\paragraph{Narrowing Types}
The procedure % $\force{v}{t}{\vsu}{\tsu}$,
\[
\forcesym : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu}
\]%
defined in Figure~\ref{fig:narrow}, takes as input a value $v$, a type
$t$, and the current value and type substitutions, and refines $v$ to
have type $t$ by yielding a triple of either the same value and
substitutions, or yields the stuck state if no such refinement is
possible. In the case where $v$ is a hole, it first checks in the given
$\vsu$ to see if the hole has already been instantiated and, if so,
returns the existing instantiation.
%
\begin{figure}[t]
\[\def\arraystretch{1.5}
\begin{array}{lll}
%% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
\forcesym                  & : v \times t \times \vsu \times \tsu \rightarrow \triple{v \cup \stuck}{\vsu}{\tsu} \\
% \force{v}{\thole}{\vsu}{\tsu}  & \defeq & \hspace{-1ex}
% \begin{cases}
%   \triple{v}{\vsu}{\tsu'} & \mbox{if } \tsu' = \unify{\{\typeof{v}, \thole\}}{\tsu} \\
%   \triple{\stuck}{\vsu}{\tsu} & \mbox{otherwise} \\

%   % \force{v}{\subst{\tsu}{\thole}}{\vsu}{\tsu} & \mbox{if}\ \thole \in dom(\tsu) \\
%   % \triple{v}{\vsu}{\extendsu{\tsu}{\thole}{\typeof{v}}} & \mbox{otherwise} \\
% \end{cases} \\
\force{\vhole{\thole}}{t}{\vsu}{\tsu} & \defeq % \hspace{-1ex}
\begin{cases}
  \triple{v}{\vsu}{\tsu'}    &\ \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}, \\
                                         \tsu' = \unify{\{\thole, t, \typeof{v}\}}{\tsu}\end{array} \\[0.16in]
  \triple{\stuck}{\vsu}{\tsu} &\ \textbf{if} \begin{array}{l} v = \lookupsu{\vsu}{\vhole{\thole}}\end{array} \\[0.08in]
  \triple{v}{\extendsu{\vsu}{\vhole{\thole}}{v}}{\tsu'} &\ \textbf{if} \begin{array}{l} \tsu' = \unify{\{\thole, t\}}{\tsu}, \\ v = \gen{t}{\tsu'} \end{array}\\
\end{cases}  \\ %[0.2in]
% \begin{cases}
%   \triple{\lookupsu{\vsu}{\ehole}}{\vsu}{\tsu} & \mbox{if}\ \ehole \in dom(\vsu), \hastype{\lookupsu{\vsu}{\ehole}}{t} \\
%   \triple{\stuck}{\vsu}{\tsu}                  & \mbox{if}\ \ehole \in dom(\vsu) \\
%   \triple{v}{\extendsu{\vsu}{\ehole}{v}}{\tsu}   & v = \gen{t}\\
% \end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \triple{\vsu(i)}{\vsu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\triple{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}{\tsu}     & \defeq \triple{n}{\vsu}{\tsu} \\
\force{b}{\tbool}{\vsu}{\tsu}    & \defeq \triple{b}{\vsu}{\tsu} \\
\force{\efun{x}{e}}{\tfun}{\vsu}{\tsu} & \defeq \triple{\efun{x}{e}}{\vsu}{\tsu} \\
\force{\epair{v_1}{v_2}}{\tprod{t_1}{t_2}}{\vsu}{\tsu} & \defeq \triple{\epair{v_1}{v_2}}{\vsu}{\tsu''} & \hspace{-43.5mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{\typeof{v_1}, t_1\}}{\tsu},\\ \tsu'' = \unify{\{\typeof{v_2}, t_2\}}{\tsu'} \end{array} \\
\force{\vleaf{t_1}}{\ttree{t_2}}{\vsu}{\tsu} & \defeq \triple{\vleaf{t_1}}{\vsu}{\tsu'} & \hspace{-43.5mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array}\\
\force{\vnode{t_1}{v_1}{v_2}{v_3}}{\ttree{t_2}}{\vsu}{\tsu} \hspace{-3mm} & \defeq \triple{\vnode{t_1}{v_1}{v_2}{v_3}}{\vsu}{\tsu'} & \hspace{-43.5mm} \textbf{if} \begin{array}{l} \tsu' = \unify{\{t_1, t_2\}}{\tsu} \end{array} \\
\force{v}{t}{\vsu}{\tsu} & \defeq \triple{\stuck}{\vsu}{\tsu}
\end{array}
\]
\caption{Narrowing values}
\label{fig:narrow}
\end{figure}
%
%While a hole may map to a value that \emph{contains} another hole, \eg a
%lambda or a tree, it may not map \emph{directly} to another hole,
As the value substitution is normalized, in the first case of \forcesym\ we
do not need to \forcesym\ the result of the substitution, the sub-hole
will be narrowed when the context demands it.

@@

\paragraph{The Type of a Value} The \emph{dynamic type}
of a value $v$ is defined as a function $\typeof{v}$ shown
in Figure~\ref{fig:typeof}.
%
The types of primitive values are defined in the natural manner.
%
The types of functions are \emph{approximated}, which is all
that is needed to ensure an application does not get stuck.
%
For example,
$$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$$
instead of $\tint \rightarrow \tint$.
%
The types of tuples are obtained directly from their values, and the
types of (polymorphic) trees from their labels.
%
Note that the evaluation relation in Figure~\ref{fig:operational}
guarantees that tree \emph{values} will be annotated with their type.
%
For nodes the type can be taken from the type of its value $v_1$, but
for leaves the evaluation relation creates a new type hole $\thole$
(this corresponds to polymorphic instantiation in a typechecker).

@@

We must address two challenges
to convert the semantics into a \emph{procedure} for finding
witnesses.
%
First, we must resolve the non-determinism introduced by \gensym.
%
Second, in the presence of higher-order functions and currying,
we must determine how many concrete values to generate to make
execution go wrong (as we cannot rely upon static typing to
provide this information.)

@@

\begin{cor}[Witness Generation]
\label{thm:generation}
  If $\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$
  then for all types $t_1 \ldots t_n$ there exist values $w_1:t_1\ \ldots\ w_n:t_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}
\begin{proof}
  For any function $f$ of multiple arguments, we can define $f'$ as the
  uncurried version of $f$ that takes all of its arguments as a single
  nested pair, and then apply Theorem~\ref{thm:soundness} to $f'$.
\end{proof}
